Inferring the distribution and rate of Earth-like exoplanets \\ from noisy individual-planet inferences

### 2014-04-19 document/ms.tex
some incomplete general discussion of hierarchical inference

% Document specific variables.
\newcommand{\rate}{\ensuremath{\Gamma}}
\newcommand{\ratepar}{{\ensuremath{\theta}}}
\newcommand{\ratepars}{{\ensuremath{\bvec{\ratepar}}}}
\newcommand{\obs}[1]{\ensuremath{\hat{#1}}}
\newcommand{\radius}{\ensuremath{R}}
\newcommand{\period}{\ensuremath{T}}
\newcommand{\completeness}{{\ensuremath{P_\mathrm{c}}}}
\newcommand{\transitprob}{{\ensuremath{P_\mathrm{t}}}}
\newcommand{\entry}{\ensuremath{w}}
\newcommand{\catalog}{{\ensuremath{\bvec{\entry}}}}

@@

\begin{abstract}
Although no true extra-solar Earth analog is known, hundreds of planets have
been found around Sun-like stars that are either Earth-sized but on shorter
periods, or else on year-long orbits but somewhat bigger than Earth.
These populations ought to permit a (possibly highly uncertain) extrapolated
measurement of the probability or rate at which Sun-like stars host true Earth
analogs.
Here we present a general probabilistic framework for making justified
inferences about the occurrence rate of exoplanets taking into account survey
completeness \emph{and observational uncertainties}.
We demonstrate that our method produces lower variance estimates than the
standard inverse-detection-efficiency weighting procedure.
When applied to an existing catalog of small planet candidates around G dwarfs
(Petigura \etal 2013), our method confirms previous measurements of a
flattening radius distribution near Earth's radius.
One output of our models is fully marginalized estimates---marginalizing out
observational uncertainties and all distribution parameters---of the rate
density of true Earth analogs, expressed as a number of planets per star per
natural logarithmic interval of period and radius, evaluated at the properties
of Earth.
The rate density we infer is an extrapolation; it depends strongly on the
permitted flexibility of the distribution model, but it tends to come out
at around ten percent, with large uncertainty.
\end{abstract}

@@

@@

\section{Maximum likelihood estimators for noiseless catalogs}

>>>>

% Document specific variables.
\newcommand{\rate}{\ensuremath{\Gamma}}
\newcommand{\ratepar}{{\ensuremath{\theta}}}
\newcommand{\ratepars}{{\ensuremath{\bvec{\ratepar}}}}
\newcommand{\obs}[1]{\ensuremath{\hat{#1}}}
\newcommand{\radius}{\ensuremath{R}}
\newcommand{\period}{\ensuremath{T}}
\newcommand{\completeness}{{\ensuremath{P_\mathrm{c}}}}
\newcommand{\transitprob}{{\ensuremath{P_\mathrm{t}}}}
\newcommand{\data}{{\ensuremath{\bvec{x}}}}
\newcommand{\entry}{{\ensuremath{\bvec{w}}}}
\newcommand{\catalog}{{\ensuremath{\bvec{\entry}}}}

@@

\begin{abstract}
Although no true extra-solar Earth analog is known, hundreds of planets have
been found around Sun-like stars that are either Earth-sized but on shorter
periods, or else on year-long orbits but somewhat bigger than Earth.
These populations ought to permit a (possibly highly uncertain) extrapolated
measurement of the probability or rate at which Sun-like stars host true Earth
analogs.
Here we present a general probabilistic framework for making justified
inferences about the occurrence rate of exoplanets taking into account survey
completeness \emph{and observational uncertainties}.
We demonstrate that our method produces lower variance estimates than the
standard inverse-detection-efficiency weighting procedure.
When applied to an existing catalog of small planet candidates around G dwarfs
(Petigura \etal\ 2013), our method confirms previous measurements of a
flattening radius distribution near Earth's radius.
One output of our models is fully marginalized estimates---marginalizing out
observational uncertainties and all distribution parameters---of the rate
density of true Earth analogs, expressed as a number of planets per star per
natural logarithmic interval of period and radius, evaluated at the properties
of Earth.
The rate density we infer is an extrapolation; it depends strongly on the
permitted flexibility of the distribution model, but it tends to come out
at around ten percent, with large uncertainty.
\end{abstract}

@@

\section{A brief introduction to hierarchical inference}

In this \paper, we are asking the question: \emph{what constraints can we put
on the occurence rate of exoplanets given all the light curves measured by
\project{Kepler}?}
Mathematically, this involves evaluating the so-called \emph{marginalized
likelihood function}
\begin{eqnarray}\eqlabel{crazylike}
p(\{\data_k\}\,|\,\ratepars) &=&
    \int\left[\prod_{k=1}^K p(\data_k\,|\,\entry_k)\right ]
    \,p(\{\entry_k\}\,|\,\ratepars)
    \dd\{\entry_k\}
\end{eqnarray}
where $\{\data_k\}$ is the set of all light curves, \ratepars\ is a vector of
parameters describing the occurence rate function $\rate_\ratepars(\entry)$
and \entry\ is a vector of physical parameters describing a planetary system
(period, radius ratio, \etc).

It is generally computationally intractable to compute the likelihood in
\eq{crazylike} because it involves the numerical evaluation of an extremely
 high dimensional---the number of physical parameters times the number of
targets---and computationally costly integral.
Instead, we can use existing catalogs of planet candidates to approximate the
marginalization integral.
For our purposes, this is the definition of \emph{hierarchical inference}.

A very simple example of such an inference is a procedure that should be very
familiar: the histogram.
In the limit of negligible uncertainties, the maximum marginalized likelihood
for the occurence rate is a histogram of the objects in the catalog (see
\app{histogram} for the derivation).

@@

\section{Maximum likelihood estimators for noiseless catalogs}
\sectlabel{histogram}
